Supplemental material of ``Microbiota: are you sick?''

### 2016-04-15 Science/science_CC_suppl.tex
Include previous releases for different publications
On branch master
	new file:   MolSysBio/CC-suppl_molsysbio.docx
	new file:   MolSysBio/molsysbio_CC.docx
	new file:   NatBiotech/CC-suppl_NatBiotech.pdf
	new file:   NatBiotech/CC-suppl_NatBiotech.tex
	new file:   NatBiotech/CC_NatBiotech.pdf
	new file:   NatBiotech/CC_NatBiotech.tex
	new file:   NatBiotech/lettertoeditor_NatBiotech.docx
	new file:   Nature/CC-suppl_sub.pdf
	new file:   Nature/CC_sub.pdf
	new file:   Nature/lettertoeditor2.docx
	new file:   Science/lettertoeditor_sub.docx
	new file:   Science/science_CC.pdf
	new file:   Scie

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

%% Template for a preprint Letter or Article for submission
%% to the journal Nature.
%% Written by Peter Czoschke, 26 February 2004
%%
\documentclass{nature} % Commented includegraphics trick in /usr/local/texlive/2014/texmf-dist/tex/latex/nature/nature.cls
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[usenames,dvipsnames,table]{xcolor}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}

\newcommand{\CC}[0]{\emph{cmplxcruncher}}
\newcommand{\task}[1]{\texttt{\bfseries\scshape\textcolor{MidnightBlue}{#1}}}
\newcommand{\un}[1]{\operatorname{#1}}
\newcommand{\unclassrate}[0]{\un{kpb/s/core}}

%% Roman footnote numbering style to not collide with bibliographic cites: 
\renewcommand{\thefootnote}{\Roman{footnote}}

%% Change floats name for supplemental material
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}

%% make sure you have the nature.cls and naturemag.bst files where
%% LaTeX can find them
\bibliographystyle{naturemag}

%%% HYPERREF
\usepackage[pdftex,
	pdftitle={Supplemental material of ``Microbiota: are you sick?''},
	pdfauthor={Jose Manuel Marti, Daniel M. Martinez, Manuel Peña, Cesar Gracia, Amparo Latorre, Andres Moya, Carlos P. Garay},
	pdfsubject={Nature Biotechnology},
	pdfkeywords={Supplemental material},
	bookmarksopen,
	colorlinks=true,
	urlcolor={red},
	bookmarks=true]{hyperref}

%% Title
\title{Supplemental material of ``Microbiota: are you sick?''}

%% Notice placement of commas and superscripts and use of &
%% in the author list
\author{Jose Manuel Mart\'i$^{1}$, Daniel M. Mart\'inez$^{1,2}$, Manuel Pe\~na$^{1}$, C\'esar Gracia$^{1}$, Amparo Latorre$^{2,3,4}$, Andr\'es Moya$^{2,3,4}$ \& Carlos P. Garay$^1$}

@@

@@

\begin{affiliations}
 \item Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain.
 \item FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain.
 \item Cavanilles Institute of Biodiversity and Evolutionary Biology, Univ. de Valencia, 46980, Spain. 
 \item CIBER en Epidemiolog\'ia y Salud P\'ublica (CIBEResp), Madrid, Spain.
\end{affiliations}

\section{Model} \label{sec:model}

@@

\subsection{Metagenomic sequences processing}
Metagenomic shotgun (and 16S too) sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT1, LMAT2} (version 1.2.4, with Feb'15 release of data base \emph{LMAT-Grand}). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It is equipped with 32 cores (64 threads available using Intel Hyper-threading technology) as it has 2 Haswell-based Xeons, the E5-2698v3@2.3 GHz, sharing half a tebibyte (0.5 TiB, that is, 512 gibibytes) of DRAM memory. This node is also provided with a card PCIe SSD as NVRAM, the P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s, which Micron kindly issued free of charge, as a sample for testing purposes. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\footnote{In this context, ``Grand''\cite{LMAT2} refers to a huge database that contains k-mers from all viral, prokaryote, fungal and protist genomes present in the NCBI database, plus Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP). This represent about 31.75 billion k-mers occupying 457.62 GB.}, release Feb'15, provided by the LMAT team\cite{LMAT2}. Previously to any calculation, the full database was loaded in the NVRAM. With this configuration the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts have been developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis.

\subsection{Taxa level selection}
We selected genus as taxonomic level for the subsequent steps of our work. To ensure that, between adjacent taxonomic levels, there were not crucial differences which could still be of relevance after standardization (see Supplementary Section \ref{sec:stan}), we tested two different datasets. In the former, the antibiotics study\cite{antibiotic} with 16S data, we tested the differences between genus and family levels. The latter dataset tested was the kwashiorkor twins study\cite{kwashiorkor} for both genus and species taxonomic levels. The Supplementary Figures \ref{fig:taxlev1} (overview) and \ref{fig:taxlev2} (detail) plot the comparison between studies (and so, 16S and SMS) and between adjacent taxonomic levels.

@@

We have studied the time dependence of the variability $V$ and power law index $\beta$ (see Supplementary Section \ref{sec:model}) by using a sliding window approach. The total number of time points are divided in subsets of five points, where next subset is defined by adding next time sampling and by eliminating the earliest one. Both parameters were calculated for each subset against the average time lapse. Supplementary Figure \ref{fig:tempevo1} shows the variability  $V$ as a function of time for the largest sampling: two individuals in the Caporaso's study\cite{moving} corresponding to the gut microbiota of a male (upper plot) and a female (lower plot). Supplementary Figure \ref{fig:tempevo2} shows the time evolution of $V$ for patient P2 of the IBS study\cite{IBS} (upper plot) and patient D in the antibiotics study\cite{antibiotic} (lower plot).

@@

The external mandatory dependencies of \CC\ are with Python packages \emph{pandas}, \emph{matplotlib}, \emph{numpy}, \emph{scipy} and \emph{xlrd}, while the recommended ones are \emph{pylatex} and \emph{uncertainties}. The PEP8 compliant code of \CC\ could be executed under a Python 2.7 interpret, but a 3.X interpret is strongly advised. The complete distribution of \CC\ has the code signature shown in Supplementary Table \ref{tab:cloc}, obtained through the \emph{CLOC}\footnote{\emph{Count Lines of Code} is a perl script that counts blank lines, comment lines, and physical lines of source code.} tool release 1.62. Supplementary Figure \ref{fig:CC} shows the code layout of the complete distribution of \CC.

@@

        \item If $\left|\mathrm{AIC_{c_{NLR}}} - \mathrm{AIC_{c_{LLR}}}\right| \leq 2$, no model is favoured, so proceed with model averaging:

@@

The colour code of the RSI percentage text in the rank plot of \CC\ is chosen following the first condition satisfied from those shown in Supplementary Table \ref{tab:RSI} (see page \pageref{tab:RSI}).

@@

\subsection{Fit Plots}
The subdirectory \texttt{fits} under the results directory contains the fit plots for each data set. Both for the unweighted fit (detailed in the Supplementary Section \ref{sec:unw}) and for the X-weighted fit (detailed in the Supplementary Section \ref{sec:X-w}), two plots are generated: the former with logarithmic scale, the latter with lineal scale. Supplementary Figures \ref{fig:unwFit} and \ref{fig:X-wFit} show examples of the unweighted and X-weighted fits, respectively. Additionally, for the unweighted fit, a complete residues analysis is performed, and a 4-in-1 figure is generated as shown in Supplementary Figure \ref{fig:unwRes} (corresponding to the fit of Supplementary Fig. \ref{fig:unwFit}). Among other tests, it allows to check for normality and homoscedasticity of the residues.

@@

\subsection{Histogram Plots} 
The subdirectory \texttt{hist} under the results directory contains the histogram plots. \CC\ generates three different histogram plots:
\begin{description}
	\item[$-$Absolute frequencies plot]: This plot is useful to visually assess the validity of the time points in terms of the accumulated absolute frequency of the elements (taxa), since absolute frequencies far (much higher or much lower) from those typically observed could mean a sampling problem. As an example, Supplementary Figure \ref{fig:histAFP} shows this histogram for the pre-treatment data (first 7 times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$2D deviation plot]: The 2D semi-logarithmic histogram representing deviations from the mean versus the mean itself, is a useful tool in the analysis of the stability of ranking processes in complex systems\cite{ranking}. Supplementary Figure \ref{fig:hist2D} shows this plot for the data used in the fit shown in Supplementary Figure \ref{fig:unwFit}.
	\item[$-$Zero relative frequency plot]: We could define the ZRF (Zero Relative Frequency, thereby ranging from 0 to 1) of an element (taxon) as the portion of times where it is zero, i.e., it is not found. Attending to all the elements (taxa), we can plot the ZRF histogram, which then lies on the horizontal axis of the plot. The vertical axis shows the number of elements (taxa), so the height of a bar represents the number of elements that have determinate ZRF. In this respect, the bar over $0.0$ counts the number of elements (taxa) that are present at every time point of the data set (aka ``core''), while the bar over $1.0$ would count the number of elements (taxa) that are never found (this bar never appears because all these ``null'' elements are automatically filtered by the code). Supplementary Figure \ref{fig:histZRF} shows an example of this plot. There, we can see that $12$ taxa are present at all the time points of the time series while $9$ taxa basically appear only once. So, this plot is clearly useful to notice how the ``core'' is distributed.
\end{description}

@@

\subsection{Correlation and Rank Plots} 
The subdirectory \texttt{corrank} under the results directory contains the correlation matrix and rank\&stability plots, as well as Excel files with the resulting matrices. \CC\ generates three different plots falling under this category:
\begin{description}
	\item[$-$Elements correlation matrix]: This plot shows a correlation matrix among the elements (taxa), calculated with the time as independent variable. For these calculations, the data set is not normalized to avoid entering an additional constraint. As an example, Supplementary Figure \ref{fig:corrElm} shows this matrix for the ``core'' elements (taxa) present in the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Times correlation matrix]: This plot presents a correlation matrix among the time points of the data set, calculated with the elements (taxa) as independent variable. Again, the data set is not normalized. Supplementary Figure \ref{fig:corrTim} shows this matrix for the ``core'' elements (taxa) present in pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Rank dynamics and stability plot]: This plot shows the variation in the rank with time for the most dominant elements (taxa) and their calculated RSI, as discussed in Supplementary Section \ref{sec:RSI}. Supplementary Figure \ref{fig:corrank} shows this plot for the elements (taxa) in the data used in the fit shown in Supplementary Figure \ref{fig:unwFit}.
\end{description}

@@

\subsection{\CC\ advanced options}
The advanced option flags are summarized here:
\vspace{-5mm} %%% CAUTION: FORCED LAYOUT
\begin{description}
	\item[$-$ZRF filter]: This is activated with \texttt{-c}/\texttt{--core} and allows global and detached results for ``core'', ``tailless'' and ``tail'', which allows comparative analysis depending on the presence/absence rate of the taxa throughout time, which is calculated in practice using the Zero Relative Frequency (ZRF) defined above. The criteria is as follows: 
	\begin{itemize}
		\item In any case, if $\operatorname{ZRF} < 0.2$ taxa are classified as ``core''. In other words, the taxa considered ``core'' are those appearing at over $80\%$ of the time points under analysis.
		\item Depending on the number of times:
		\begin{itemize}
			\item If there are only 3 time points, taxa with $0.2 \le\operatorname{ZRF} < 0.4$ are classified as ``tailless'' and, therefore, taxa with $\operatorname{ZRF} \ge 0.4$ are classified as ``tail''.
			\item If there are 4 time points, taxa with $0.2 \le\operatorname{ZRF} < 0.5$ are classified as ``tailless'' and, therefore, taxa with $\operatorname{ZRF} \ge 0.5$ are classified as ``tail''.
			\item If there are 5 or more time points, taxa with $0.2 \le\operatorname{ZRF} < 0.6$ are classified as ``tailless'' and, therefore, taxa with $\operatorname{ZRF} \ge 0.6$ are classified as ``tail''.			      
		\end{itemize}
    \end{itemize}
	\item[$-$Relative frequency filters]: The minimum frequency filter is activated with \texttt{-z}/\texttt{--fmin} and the maximum one with \texttt{-x}/\texttt{--fmax}, with values between $0$ and $1$. If both filters are activated simultaneously, the minimum frequency cannot be equal or higher than the maximum one. If no filter is activated, by default, \CC\ is filtering all the taxa that never appear throughout time, in the event there are such taxa among the input data. On the one hand, if a minimum frequency value is provided, all the taxa whose relative accumulated frequencies (the normalized sum of the values through time) are under such a value, are eliminated. But in addition, once the set of data is normalized (to work with relative frequencies), if there are data at any time points that are under the minimum frequency given value, they are automatically floored to zero (we call this \emph{item@time filtering}), and the set of data is renormalized. On the other hand, if a maximum frequency value is provided, all the taxa whose relative accumulated frequencies are over such value, are deleted. 
\end{description}

\section{\CC\ input}
The input data are provided by two simultaneously available ways:
\vspace{-5mm} %%% CAUTION: FORCED LAYOUT
\begin{enumerate}
	\item Text files (\texttt{txt} extension), to be processed in parallel\footnote{Currently, parallelization in \CC\ is only available under GNU/Linux}, with a shared-memory model.
	\item Excel files (\texttt{xls[x]} extension), whose sheets are related datasets to be processed in parallel\footnotemark[\value{footnote}]. Optionally, all the sheets of an Excel file are standardized using part of such Excel datasets, following the procedure described in the Supplementary Section \ref{sec:stan}, in which event, additional output is provided by \CC.
\end{enumerate}
Typically, the data sets are tables with the taxa lying in different rows and the time points ordered in different columns. So, the first column is considered the taxon labels, whereas the first row contains the time labels. It is worth mentioning that any Excel sheet or text file beginning with the underscore character is ignored by \CC, as they are supposed to hold metadata.

@@

For a Taylor parameter, e.g. $V$, the estimate of the mean ($\widehat{V}$) for the healthy subpopulation, composed of $h$ individuals, is:
$$\widehat{V} = \frac{1}{W_1}\sum_{i=1}^h V_i \omega_i=\sum_{i=1}^h V_i \omega_i$$
as $W_1=\sum_i^h \omega_i=1$, since $\omega_i$ are normalized weights calculated as:
$$\omega_i = \frac{\frac{1}{\sigma^2_{V_i}}}{\sum_i^h\frac{1}{\sigma^2_{V_i}}}$$
being $\sigma_{V_i}$ the estimation of the uncertainty in $V_i$ obtained together with $V_i$ from the X-weighted power-law fit described in Supplementary Section \ref{sec:X-w}, for healthy individuals.

@@

\section{\CC\ interface with LMAT} \label{sec:lmat}
With the aim of adapting LMAT (Livermore Metagenomics Analysis Toolkit) software\cite{LMAT1, LMAT2} for its easy use in time series data, like the ones from the Kwashiorkor paper\cite{kwashiorkor}, we developed Python front and back-end scripts, which are supplied in the directory \texttt{tools} of the \CC\ package.

>>>>

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

% Users of the {thebibliography} environment or BibTeX should use the
% scicite.sty package, downloadable from *Science* at
% www.sciencemag.org/about/authors/prep/TeX_help/ .
% This package should properly format in-text
% reference calls and reference-list numbers.

\usepackage{scicite}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[usenames,dvipsnames,table]{xcolor}
\usepackage[nolists,tablesfirst, nomarkers]{endfloat}

% Use times if you have the font installed; otherwise, comment out the
% following line.

\usepackage{times}

% The preamble here sets up a lot of new/revised commands and
% environments.  It's annoying, but please do *not* try to strip these
% out into a separate .sty file (which could lead to the loss of some
% information when we convert the file to other formats).  Instead, keep
% them in the preamble of your main LaTeX source file.

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

%The next command sets up an environment for the abstract to your paper.

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

% If your reference list includes text notes as well as references,
% include the following line; otherwise, comment it out.

\newcommand{\CC}[0]{\emph{cmplxcruncher}}
\newcommand{\task}[1]{\texttt{\bfseries\scshape\textcolor{MidnightBlue}{#1}}}
\newcommand{\un}[1]{\operatorname{#1}}
\newcommand{\unclassrate}[0]{\un{kpb/s/core}}
\renewcommand\refname{References and Notes}
\renewcommand{\thefootnote}{\Roman{footnote}}

% The following lines set up an environment for the last note in the
% reference list, which commonly includes acknowledgments of funding,
% help, etc.  It's intended for users of BibTeX or the {thebibliography}
% environment.  Users who are hand-coding their references at the end
% using a list environment such as {enumerate} can simply add another
% item at the end, and it will be numbered automatically.

\newcounter{lastnote}
\newenvironment{scilastnote}{%
\setcounter{lastnote}{\value{enumiv}}%
\addtocounter{lastnote}{+1}%
\begin{list}%
{\arabic{lastnote}.}
{\setlength{\leftmargin}{.22in}}
{\setlength{\labelsep}{.5em}}}
{\end{list}}

% Include your paper's title here

\title{Supplemental material of ``Microbiota: are you sick?''}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{Jose Manuel Mart\'i,$^{1}$ Daniel M. Mart\'inez,$^{2}$ Manuel Pe\~na$^{1}$, C\'esar Gracia$^{1}$, \\
Amparo Latorre$^{2,3,4}$, Andr\'es Moya$^{2,3,4}$ \& Carlos P. Garay$^{1,\ast}$\\
\\
\normalsize{$^{1}$Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain.}\\
\normalsize{$^{2}$FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain.}\\
\normalsize{$^{3}$Cavanilles Institute of Biodiversity and Evolutionary Biology, Univ. de Valencia, 46980, Spain.}\\
\normalsize{$^{4}$CIBER en Epidemiolog\' ia y Salud P\'ublica (CIBEResp), Madrid, Spain.}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  penya@ific.uv.es}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

@@

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

@@

% In setting up this template for *Science* papers, we've used both
% the \section* command and the \paragraph* command for topical
% divisions.  Which you use will of course depend on the type of paper
% you're writing.  Review Articles tend to have displayed headings, for
% which \section* is more appropriate; Research Articles, when they have
% formal topical divisions at all, tend to signal them with bold text
% that runs into the paragraph, for which \paragraph* is the right
% choice.  Either way, use the asterisk (*) modifier, as shown, to
% suppress numbering.

\section*{Model} \label{sec:model}

@@

\subsection{Metagenomic sequences processing}
Metagenomic shotgun (and 16S too) sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of data base \emph{LMAT-Grand}). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It is equipped with 32 cores (64 threads available using Intel Hyper-threading technology) as it has 2 Haswell-based Xeons, the E5-2698v3@2.3 GHz, sharing half a tebibyte (0.5 TiB, that is, 512 gibibytes) of DRAM memory. This node is also provided with a card PCIe SSD as NVRAM, the P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s, which Micron kindly issued free of charge, as a sample for testing purposes. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\footnote{In this context, ``Grand'' refers to a huge database that contains k-mers from all viral, prokaryote, fungal and protist genomes present in the NCBI database, plus Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP). This represent about 31.75 billion k-mers occupying 457.62 GB.}, release Feb'15, provided by the LMAT team. Previously to any calculation, the full database was loaded in the NVRAM. With this configuration the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts have been developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis.

\subsection{Taxa level selection}
We selected genus as taxonomic level for the subsequent steps of our work. In order to ensure that, between adjacent taxonomic levels, there were not crucial differences which could still be of relevance after standardization (see Section \ref{sec:stan}), we tested two different data sets. In the former, the antibiotics study\cite{antibiotic} with 16S data, we tested the differences between genus and family levels. The latter dataset tested was the kwashiorkor discordant twins study\cite{kwashiorkor} for both genus and species taxonomic levels. The Supplementary Figures \ref{fig:taxlev1} (overview) and \ref{fig:taxlev2} (detail) plot the comparison between studies (and so, 16S and SMS) and between adjacent taxonomic levels.

@@

We have studied the time dependence of the variability $V$ and power law index $\beta$ (see Section \ref{sec:model}) by using a sliding window approach. The total number of time points are divided in subsets of five points, where next subset is defined by adding next time sampling and by eliminating the earliest one. Both parameters were calculated for each subset against the average time lapse. Figure \ref{fig:tempevo1} shows the variability  $V$ as a function of time for the largest sampling: two individuals in the Caporaso's study\cite{moving} corresponding to the gut microbiota of a male (upper plot) and a female (lower plot). Figure \ref{fig:tempevo2} shows the time evolution of $V$ for patient P2 of the IBS study\cite{IBS} (upper plot) and patient D in the antibiotics study\cite{antibiotic} (lower plot).

@@

The external mandatory dependencies of \CC\ are with Python packages \emph{pandas}, \emph{matplotlib}, \emph{numpy}, \emph{scipy} and \emph{xlrd}, while the recommended ones are \emph{pylatex} and \emph{uncertainties}. The PEP8 compliant code of \CC\ could be executed under a Python 2.7 interpret, but a 3.X interpret is strongly advised. The complete distribution of \CC\ has the code signature shown in the Table \ref{tab:cloc}, obtained through the \emph{CLOC}\footnote{\emph{Count Lines of Code} is a perl script that counts blank lines, comment lines, and physical lines of source code.} tool release 1.62. Figure \ref{fig:CC} shows the code layout of the complete distribution of \CC.

@@

        \item If $\left|\mathrm{AIC_{c_{NLR}}} - \mathrm{AIC_{c_{LLR}}}\right| \leq 2$, no model is favoured, so proceed with averaging:

@@

The colour code of the RSI percentage text in the rank plot of \CC\ is chosen following the first condition satisfied from those shown in Table \ref{tab:RSI} (see page \pageref{tab:RSI}).

@@

\subsection{Fit Plots}
The subdirectory \texttt{fits} under the results directory contains the fit plots for each data set. Both for the unweighted fit (detailed in the section \ref{sec:unw}) and for the X-weighted fit (detailed in the section \ref{sec:X-w}), two plots are generated: the former with logarithmic scale, the latter with lineal scale. Figure \ref{fig:unwFit} shows an example of the unweighted fit plots and Figure \ref{fig:X-wFit} does the same with the X-weighted fit plots. Additionally, for the unweighted fit, a complete residues analysis is performed, and a 4-in-1 figure is generated as shown in Figure \ref{fig:unwRes}, corresponding to the fit of Figure \ref{fig:unwFit}. Among other tests, it allows to check for normality and homoscedasticity of the residues.

@@

\subsection{Histogram Plots} 
The subdirectory \texttt{hist} under the results directory contains the histogram plots. \CC\ generates three different histogram plots:
\begin{description}
	\item[$-$Absolute frequencies plot]: This plot is useful to visually assess the validity of the time points in terms of the accumulated absolute frequency of the elements (taxa), since absolute frequencies far (much higher or much lower) from those typically observed could mean a sampling problem. As an example, Figure \ref{fig:histAFP} shows this histogram for the pre-treatment data (first 7 times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$2D deviation plot]: The 2D semi-logarithmic histogram representing deviations from the mean versus the mean itself, is a useful tool in the analysis of the stability of ranking processes in complex systems\cite{ranking}. Figure \ref{fig:hist2D} shows this plot for the data used in the fit shown in Figure \ref{fig:unwFit}.
	\item[$-$Zero relative frequency plot]: We could define the ZRF (Zero Relative Frequency, thereby ranging from 0 to 1) of an element (taxon) as the portion of times where it is zero, i.e., it is not found. Attending to all the elements (taxa), we can plot the ZRF histogram, which then lies on the horizontal axis of the plot. The vertical axis shows the number of elements (taxa), so the height of a bar represents the number of elements that have determinate ZRF. In this respect, the bar over $0.0$ counts the number of elements (taxa) that are present at every time point of the data set (aka ``core''), while the bar over $1.0$ would count the number of elements (taxa) that are never found (this bar never appears because all these ``null'' elements are automatically filtered by the code). Figure \ref{fig:histZRF} shows an example of this plot. There, we can see that $12$ taxa are present at all the time points of the time series while $9$ taxa basically appear only once. So, this plot is clearly useful to notice how the ``core'' is distributed.
\end{description}

@@

\subsection{Correlation and Rank Plots} 
The subdirectory \texttt{corrank} under the results directory contains the correlation matrix and rank\&stability plots, as well as Excel files with the resulting matrices. \CC\ generates three different plots falling under this category:
\begin{description}
	\item[$-$Elements correlation matrix]: This plot shows a correlation matrix among the elements (taxa), calculated with the time as independent variable. For these calculations, the data set is not normalized to avoid entering an additional constraint. As an example, Figure \ref{fig:corrElm} shows this matrix for the ``core'' elements (taxa) present in the pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Times correlation matrix]: This plot presents a correlation matrix among the time points of the data set, calculated with the elements (taxa) as independent variable. Again, the data set is not normalized. Figure \ref{fig:corrTim} shows this matrix for the ``core'' elements (taxa) present in pre-treatment data (first seven times) of patient ``D'' in the antibiotics study\cite{antibiotic}. 
	\item[$-$Rank dynamics and stability plot]: This plot shows the variation in the rank with time for the most dominant elements (taxa) and their calculated RSI, as discussed in Section \ref{sec:RSI}. Figure \ref{fig:corrank} shows this plot for the elements (taxa) in the data used in the fit shown in Figure \ref{fig:unwFit}.
\end{description}

@@

\subsection{\CC\ advanced options}
The advanced option flags are summarized here:
\begin{description}
	\item[$-$ZRF filter]: This is activated with \texttt{-c}/\texttt{--core} and allows global and detached results for ``core'', ``tailless'' and ``tail'', which allows comparative analysis depending on the presence/absence rate of the taxa throughout time, which is calculated in practice using the Zero Relative Frequency (ZRF) defined above. The criteria is as follows: 
	\begin{itemize}
		\item In any case, if $\operatorname{ZRF} < 0.2$ taxa are classified as ``core''. In other words, the taxa considered ``core'' are those appearing at over $80\%$ of the time points under analysis.
		\item Depending on the number of times:
		\begin{itemize}
			\item If there are only 3 time points, taxa with $0.2 \le\operatorname{ZRF} < 0.4$ are classified as ``tailless'' and, therefore, taxa with $\operatorname{ZRF} \ge 0.4$ are classified as ``tail''.
			\item If there are 4 time points, taxa with $0.2 \le\operatorname{ZRF} < 0.5$ are classified as ``tailless'' and, therefore, taxa with $\operatorname{ZRF} \ge 0.5$ are classified as ``tail''.
			\item If there are 5 or more time points, taxa with $0.2 \le\operatorname{ZRF} < 0.6$ are classified as ``tailless'' and, therefore, taxa with $\operatorname{ZRF} \ge 0.6$ are classified as ``tail''.			      
		\end{itemize}
    \end{itemize}
	\item[$-$Relative frequency filters]: The minimum frequency filter is activated with \texttt{-z}/\texttt{--fmin} and the maximum one with \texttt{-x}/\texttt{--fmax}, with values between $0$ and $1$. If both filters are activated simultaneously, the minimum frequency cannot be equal or higher than the maximum one. If no filter is activated, by default, \CC\ is filtering all the taxa that never appear throughout time, in the event there are such taxa among the input data. On the one hand, if a minimum frequency value is provided, all the taxa whose relative accumulated frequencies (the normalized sum of the values through time) are under such a value, are eliminated. But in addition, once the set of data is normalized (to work with relative frequencies), if there are data at any time points that are under the minimum frequency given value, they are automatically floored to zero (we call this \emph{item@time filtering}), and the set of data is renormalized. On the other hand, if a maximum frequency value is provided, all the taxa whose relative accumulated frequencies are over such value, are deleted. 
\end{description}

\section{\CC\ input}
The input data are provided by two simultaneously available ways:
\begin{enumerate}
	\item Text files (\texttt{txt} extension), to be processed in parallel\footnote{Currently, parallelization in \CC\ is only available under GNU/Linux}, with a shared-memory model.
	\item Excel files (\texttt{xls[x]} extension), whose sheets are related datasets to be processed in parallel\footnotemark[\value{footnote}]. Optionally, all the sheets of an Excel file are standardized using part of such Excel datasets, following the procedure described in the Supplementary Section \ref{sec:stan}, in which event, additional output is provided by \CC.
\end{enumerate}
Typically, the data sets are tables with the taxa lying in different rows and the time points ordered in different columns. So, the first column is considered the taxon labels, whereas the first row contains the time labels. It is worth mentioning that any Excel sheet or text file beginning with the underscore character is ignored by \CC, as they are supposed to hold metadata.

@@

For a Taylor parameter, e.g. $V$, the estimate of the mean ($\widehat{V}$) for the healthy subpopulation, composed of $h$ individuals, is:
$$\widehat{V} = \frac{1}{W_1}\sum_{i=1}^h V_i \omega_i=\sum_{i=1}^h V_i \omega_i$$
as $W_1=\sum_i^h \omega_i=1$, since $\omega_i$ are normalized weights calculated as:
$$\omega_i = \frac{\frac{1}{\sigma^2_{V_i}}}{\sum_i^h\frac{1}{\sigma^2_{V_i}}}$$
being $\sigma_{V_i}$ the estimation of the uncertainty in $V_i$ obtained together with $V_i$ from the X-weighted power-law fit described in Section \ref{sec:X-w}, for healthy individuals.

@@

\section{\CC\ interface with LMAT} \label{sec:lmat}
With the aim of adapting LMAT (Livermore Metagenomics Analysis Toolkit) software\cite{LMAT} for its easy use in time series data, like the ones we have analyzed in this paper\cite{kwashiorkor}, we developed Python front and back-end scripts. They are supplied in the directory \texttt{tools} of the \CC\ package.
